---
title: "Logistic regression"
output: html_document
date: "2024-05-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r include=FALSE}
#libraries
#install.packages(c("tidyverse","ggplot2","psych","rstatix","readxl"))
library("tidyverse")
library("ggplot2")
#install.packages("kernlab")
library(kernlab)
#install.packages("ggcorrplot")
library(ggcorrplot)
library(psych)
#install.packages("ltm")
library(ltm)
library(rstatix)
library(readxl)
library(dplyr)
```


## Logistic regression

I used data_final_mortality.csv
Coffee: Patient consumed coffee (1) or not (0)
CaffeinatedStatus: Caffeinated Coffee(1) | Decaffeited Coffee(0)
mortality: morstat: 0-alive, 1- dead

```{r echo=FALSE}
data <- read.csv("df_final_mortality.csv")
summary(data)
data <- data%>% 
  rename(Age_num = RIDAGEYR,
         Sex = RIAGENDR,
         Education = DMDEDUC,
         Race = RIDRETH1,
         MaritalStatus = DMDMARTL,
         PovertyIncome = INDFMPIR) %>%
  mutate(Sex = as.factor(Sex),
         Education = as.factor(Education),
         Race = as.factor(Race),
         MaritalStatus = as.factor(MaritalStatus),
         PAD = as.factor(PAD),
         Coffee = as.factor(Coffee),
         CaffeinatedStatus = as.factor(CaffeinatedStatus),
         SugaryStatus = as.factor(SugaryStatus),
         FattyStatus = as.factor(FattyStatus),
         MilkContainingStatus = as.factor(MilkContainingStatus)) %>%
  mutate(Sex = plyr::revalue(Sex, c("1" = "Male","2" = "Female")),
         Education = case_when(Education == "1" ~ "Less Than High School",
                               Education == "2" ~ "High School",
                               Education == "3" ~ "Some college or above",
                               Education == "7" | Education == "9" ~ "Incomplete"), #patients REFUSED to give into about the education or they DON'T KNOW
         Education = as.factor(Education),
         MaritalStatus = case_when(MaritalStatus == "1" ~ "Married",
                                   MaritalStatus == "5" ~ "Never Married",
                                   MaritalStatus =="2" | MaritalStatus == "3" | MaritalStatus == "4" | MaritalStatus == "6" ~ "Unmarried but have or had a partner"),
         MaritalStatus = as.factor(MaritalStatus),
         Race = case_when(Race == "1" ~ "Mexican American",
                          Race == "4" ~ "Non-Hispanic Black",
                          Race == "3" ~ "Non-Hispanic White",
                          Race == "2" | Race == "5" ~ "Other"),
         Race = as.factor(Race),
         Age_fct = case_when(Age_num < 65 ~ "Age < 65 years",
                             Age_num >= 65 ~ "Age >= 65 years"),
         Age_fct = as.factor(Age_fct)) %>%
  filter(LEXLABPI < 1.40 & LEXRABPI <1.40) 
```

# Coffee consumption (Age, Sex)


```{r  echo=FALSE}
model5 <- glm(mortstat ~ Coffee, family = binomial(), data = data)
summary(model5)

```
            Estimate Std. Error z value Pr(>|z|)    
(Intercept)   0.6419     0.1747   3.674 *0.000239* ***
Coffee1       0.7007     0.2171   3.228 *0.001246* ** 

The results indicate that coffee consumption (Coffee1) is significantly associated with higher mortality rates in the sample. Specifically, the odds of dying for patients who consumed coffee are exp(0.7007) times higher than for those who did not consume coffee.
The statistical significance (p < 0.05) suggests that the observed association between coffee consumption and mortality is not likely due to random chance in this sample.

Intercept (0.6419): This value represents the log odds of mortality for patients who did not consume coffee, when all other predictors in the model are held constant (in this case, coffee consumption is the only predictor, so this simply represents non-coffee drinkers). To convert this log odds into odds:
This means that the odds of mortality for non-coffee drinkers are about 1.9, or they are 1.9 times more likely to die than not to die.

Coffee Coefficient (0.7007): This coefficient indicates the change in log odds due to consuming coffee. The odds ratio can be calculated by exponentiating this coefficient:

This odds ratio means that coffee drinkers have about 2.02 times the odds of dying compared to non-coffee drinkers.

```{r  echo=FALSE}


model6 <- glm(mortstat ~ Coffee + Age_num + Sex, family = binomial(), data = data)
summary(model6)


```
Here coffee variable was not statistically significant compared to other variables in the model. Age is a huge factor, as well as being female (probability of dying is lower)

```{r  echo=FALSE}

model7 <- glm(mortstat ~ Coffee + Age_num, family = binomial(), data = data)
summary(model7)


```
As well, coffee is not statistically significant.


```{r  echo=FALSE}

model8 <- glm(mortstat ~ Coffee + Sex, family = binomial(), data = data)
summary(model8)

```

This coefficient suggests that consuming coffee is associated with an increase in the log odds of mortality by 0.5925 when compared to not consuming coffee, controlling for sex.
The p-value of 0.0082 indicates that this effect is statistically significant, implying a relationship between coffee consumption and increased mortality risk in the context of this model.
SexFemale (-1.1405):

This coefficient indicates that being female is associated with a reduction in the log odds of mortality by 1.1405 compared to males, controlling for coffee consumption.
The coefficient is significantly different from zero (p = 3.12e-07), suggesting a strong protective effect of being female on the risk of mortality within this dataset.

## Decaf versus Caf 

```{r}
model9 <- glm(mortstat ~ CaffeinatedStatus, family = binomial(), data = data)
summary(model9)


```
The model suggests that consuming caffeinated coffee is associated with a statistically significant decrease in the risk of mortality compared to not consuming caffeinated coffee, controlling for other factors not included in this model. However, given the closeness of the p-value to 0.05 and the presence of missing data, further investigation could be beneficial, possibly exploring additional variables that could affect or modify this relationship.

```{r}
model10 <- glm(mortstat ~ CaffeinatedStatus + Age_num + Sex, family = binomial(), data = data)
summary(model10)


```
not significant 


```{r}
model11 <- glm(mortstat ~ CaffeinatedStatus + Age_num, family = binomial(), data = data)
summary(model11)

```
not significant

```{r}

model12 <- glm(mortstat ~ CaffeinatedStatus + Sex, family = binomial(), data = data)
summary(model12)
```
Drinking caffeinated coffee reduced the risk of mortality. 

# Age factor and other factors

```{r}
model20 <- glm(mortstat ~ Age_fct, family = binomial(), data = data)
summary(model20)
```

Age is significant factor..

```{r}
model24 <- glm(mortstat ~ Age_fct + Coffee, family = binomial(), data = data)
summary(model24)


```
Interesting, we see that drinking coffee is associated with higher mortality rates in case with individuals that are over 65. 
```{r}
model25 <- glm(mortstat ~ Age_fct + CaffeinatedStatus, family = binomial(), data = data)
summary(model25)
```
But caffeine inside coffee did not have any major consequences to mortality.. 
